---
title: "Practical Machine Learning Project Report"
author: "Suman Mandal"
date: "Saturday, August 06, 2016"
output: html_document
---

## Prediction Assignment

### Background
Using devices such as JawboneUp, NikeFuelBand, and Fitbitit is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in
their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it.  
   
In this project, the goal is to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website: [http://groupware.les.inf.puc-rio.br/har](http://groupware.les.inf.puc-rio.br/har) (see the section on the Weight Lifting Exercise Dataset).   

### Preparing the data and R packages  

#### Load packages, set caching 

```{r, message=FALSE}
require(caret)
require(corrplot)
require(Rtsne)
require(xgboost)
require(stats)
require(knitr)
require(ggplot2)
require(Ckmeans.1d.dp)
knitr::opts_chunk$set(cache=TRUE)
```

For fast and accurate training of the model, I have chosen XGBoost, an implementation of tree-based extreme gradient boosting algorithm. (*The XGBoost tool is confirmed by course's CTA to be allowed in this assignment.*)   

#### Getting the Data

```{r}
# URL of the training and testing data
train.url ="http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
test.url = "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
# file names
train.name = "./data/pml-training.csv"
test.name = "./data/pml-testing.csv"
# if directory does not exist, create new
if (!file.exists("./data")) {
  dir.create("./data")
}
# if files does not exist, download the files
if (!file.exists(train.name)) {
  download.file(train.url, destfile=train.name)
}
if (!file.exists(test.name)) {
  download.file(test.url, destfile=test.name)
}
# load the CSV files as data.frame 
train = read.csv("./data/pml-training.csv")
test = read.csv("./data/pml-testing.csv")
dim(train)
dim(test)
names(train)
```  

The raw training data has 19622 rows of observations and 158 features (predictors). Column `X` is unusable row number. While the testing data has 20 rows and the same 158 features. There is one column of target outcome named `classe`.   

#### Cleaning the Raw Data

At the begining extract the target outcome (the activity quality) from training data, so that now the training data only contains the predictors (the activity monitors).

```{r}
# target outcome (label)
outcome.org = train[, "classe"]
outcome = outcome.org 
levels(outcome)
```

The outcome has 5 levels in character format.   
Convert the outcome to numeric, because XGBoost gradient booster only recognizes numeric data.

```{r}
# convert character levels to numeric
num.class = length(levels(outcome))
levels(outcome) = 1:num.class
head(outcome)
```
   
The outcome is removed from training data.   
```{r}
# remove outcome from train
train$classe = NULL
```

The assignment require us to use data from accelerometers on the `belt`, `forearm`, `arm`, and `dumbell`, so these features are extracted based on these keywords.   
  
```{r}
# filter columns on: belt, forearm, arm, dumbell
filter = grepl("belt|arm|dumbell", names(train))
train = train[, filter]
test = test[, filter]
```

Instead of less-accurate imputation of missing data, now we remove all columns with NA values.   
```{r}
# remove columns with NA, use test data as referal for NA
cols.without.na = colSums(is.na(test)) == 0
train = train[, cols.without.na]
test = test[, cols.without.na]
```

### Preprocessing the data

#### Check for features's variance

Based on the principal component analysis PCA, it is important that features have maximum variance for maximum uniqueness, so that each feature is as distant as possible (as orthogonal as possible) from the other features.

```{r}
# check for zero variance
zero.var = nearZeroVar(train, saveMetrics=TRUE)
zero.var
```
There is no feature without variability (all of them have enough variance). So there is no feature to be removed further.

#### Plot the relationship between features and outcome  

Plot the relationship between features and outcomes. From the plot below, each feature has relatively the same distribution among the 5 outcome levels (A, B, C, D, E).

```{r fig.width=12, fig.height=8, dpi=72}
featurePlot(train, outcome.org, "strip")
```

#### Plot the correlation matrix  

Plot a correlation matrix between features.   
A good set of feature is when they are highly uncorrelated (orthogonal) between each other. The plot below shows average of correlation is not too high, so I have chosen not to perform further PCA preprocessing.

```{r fig.width=12, fig.height=12, dpi=72}
corrplot.mixed(cor(train), lower="circle", upper="color", 
               tl.pos="lt", diag="n", order="hclust", hclust.method="complete")
```

#### tSNE plot 

A tSNE (t-Distributed Stochastic Neighbor Embedding) visualization is a 2D plot of multi-dimensional features, that is multi-dimensional reduction into a 2D plane. In the tSNE plot below there is no clear separation of clustering of the 5 levels of outcome (A, B, C, D, E). So we cannot get to a conclusion to manually build any regression equation from the irregularity.   

```{r fig.width=12, fig.height=8, dpi=72}
# t-Distributed Stochastic Neighbor Embedding
tsne = Rtsne(as.matrix(train), check_duplicates=FALSE, pca=TRUE, 
              perplexity=30, theta=0.5, dims=2)
embedding = as.data.frame(tsne$Y)
embedding$Class = outcome.org
g = ggplot(embedding, aes(x=V1, y=V2, color=Class)) +
  geom_point(size=1.25) +
  guides(colour=guide_legend(override.aes=list(size=6))) +
  xlab("") + ylab("") +
  ggtitle("t-SNE 2D Embedding of 'Classe' Outcome") +
  theme_light(base_size=20) +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank())
print(g)
```

### Build the Machine Learning model 

Now building a Machine Learning model to predict activity quality (`classe` outcome) from the activity monitors (the features or predictors) by using XGBoost extreme gradient boosting algorithm.

#### XGBoost data

XGBoost supports only numeric matrix data. Converting all training, testing and outcome data to matrix.

```{r}
# convert data to matrix
train.matrix = as.matrix(train)
mode(train.matrix) = "numeric"
test.matrix = as.matrix(test)
mode(test.matrix) = "numeric"
# convert outcome from factor to numeric matrix 
#   xgboost takes multi-labels in [0, numOfClass)
y = as.matrix(as.integer(outcome)-1)
```

#### XGBoost parameters 

Set XGBoost parameters for cross validation and training.  
Set a multiclass classification objective as the gradient boosting's learning function.   
Set evaluation metric to `merror`, multiclass error rate.   

```{r}
# xgboost parameters
param <- list("objective" = "multi:softprob",    # multiclass classification 
              "num_class" = num.class,    # number of classes 
              "eval_metric" = "merror",    # evaluation metric 
              "nthread" = 8,   # number of threads to be used 
              "max_depth" = 16,    # maximum depth of tree 
              "eta" = 0.3,    # step size shrinkage 
              "gamma" = 0,    # minimum loss reduction 
              "subsample" = 1,    # part of data instances to grow tree 
              "colsample_bytree" = 1,  # subsample ratio of columns when constructing each tree 
              "min_child_weight" = 12  # minimum sum of instance weight needed in a child 
              )
```

#### Expected Error Rate 

Expected error rate is less than `1%` for a good classification. Do a cross validation to estimate the error rate using 4-fold cross validation, with 200 epochs to reach the expected error rate of less than `1%`.

#### 4-fold cross validation  

```{r}
# set random seed, for reproducibility 
set.seed(1234)
# k-fold cross validation, with timing
nround.cv = 200
system.time( bst.cv <- xgb.cv(param=param, data=train.matrix, label=y, 
              nfold=4, nrounds=nround.cv, prediction=TRUE, verbose=FALSE) )
```

Elapsed time is around 150 seconds (2.5 minutes).  

```{r}
tail(bst.cv$dt) 
```

From the cross validation set, choose index with minimum multiclass error rate.  
Index will be used in the model training to fulfill expected minimum error rate of `< 1%`.

```{r}
# index of minimum merror
min.merror.idx = which.min(bst.cv$dt[, test.merror.mean]) 
min.merror.idx 
# minimum merror
bst.cv$dt[min.merror.idx,]
```

Best cross-validation's minimum error rate `test.merror.mean` is around 0.006 (0.6%). It happened at 106th iteration.

#### The Confusion Matrix 

It tabulates the cross-validation's predictions of the model against the truths.

```{r}
# get CV's prediction decoding
pred.cv = matrix(bst.cv$pred, nrow=length(bst.cv$pred)/num.class, ncol=num.class)
pred.cv = max.col(pred.cv, "last")
# confusion matrix
confusionMatrix(factor(y+1), factor(pred.cv))
```

As is expected the Confusion matrix shows concentration of correct predictions is in the diagonal.  
The average accuracy is `99.38%`, with error rate `0.62%`. So, expected error rate of less than `1%` is fulfilled.

#### Model Training 

Fit the XGBoost gradient boosting model on all of the training data.   

```{r}
# real model fit training, with full data
system.time( bst <- xgboost(param=param, data=train.matrix, label=y, 
                           nrounds=min.merror.idx, verbose=0) )
```
Time elapsed is around 35 seconds.  

#### Predicting the testing data

```{r}
# xgboost predict test data using the trained model
pred <- predict(bst, test.matrix)  
head(pred, 10)  
```

#### Post Processing

The Output of prediction is the predicted probability of the 5 levels (columns) of outcome.  
Decode the quantitative 5 levels of outcomes to qualitative letters (A, B, C, D, E).   
  
```{r}
# decode prediction
pred = matrix(pred, nrow=num.class, ncol=length(pred)/num.class)
pred = t(pred)
pred = max.col(pred, "last")
pred.char = toupper(letters[pred])
```

(*The prediction result `pred.char` is not displayed intentionally due to Honour Code, because it is the answer of the "project submission" part.*)

#### Feature Importance

```{r fig.width=8, fig.height=12, dpi=72}
# get the trained model
model = xgb.dump(bst, with.stats=TRUE)
# get the feature real names
names = dimnames(train.matrix)[[2]]
# compute feature importance matrix
importance_matrix = xgb.importance(names, model=bst)

# plot
gp = xgb.plot.importance(importance_matrix)
print(gp) 
```

Feature Importance plot is useful to select only best features with highest correlation to the outcome(s). To improve model fitting performance (time or overfitting), less important features can be removed.

### Creating the Submission Files 

```{r}
# if directory does not exist, create new
if (!file.exists("./answer")) {
  dir.create("./answer")
}
path = "./answer"
pml_write_files = function(x) {
    n = length(x)
    for(i in 1: n) {
        filename = paste0("problem_id_", i, ".txt")
        write.table(x[i], file=file.path(path, filename), 
                    quote=FALSE, row.names=FALSE, col.names=FALSE)
    }
}
pml_write_files(pred.char)
```
------------------